An Improved Method for Coupling Hydrodynamics with Astrophysical Reaction Networks

### 2022-05-09 simplified_sdc_intro/paper.tex
more work

    \item construct the advective update terms, $\Advt{\Uc}^{n+1/2,(k)}_{i}$,
      by first ignoring the hydrodynamics sources,
      \begin{equation}
        \Advt{\Uc}^{n+1/2,(k)}_{i} =
          - \frac{\Fb^\xv(\Uc^{n+1/2,(k)}_{i+1/2}) - \Fb^\xv(\Uc^{n+1/2,(k)}_{i-1/2})}{\Delta x}
      \end{equation}
      for the general case, and
      \begin{align}
        \Advsst{\rho e}^{n+1/2,(k)}_{i} =
          &- \frac{F\left((\rho e)^{n+1/2,(k)}_{i+1/2}\right) - F\left((\rho e)^{n+1/2,(k)}_{i-1/2}\right)}{\Delta x} \\
          &- \left (\frac{p_{i+1/2}^{n+1/2,(k)} + p_{i-1/2}^{n+1/2,(k)}}{2} \right )\frac{u_{i+1/2}^{n+1/2,(k)} - u_{i-1/2}^{n+1/2,(k)}}{\Delta x}
      \end{align}
      for $(\rho e)$.

@@

The initial model is generated following the procedure described in
\citep{subchandra}.  We use a $1.1~M_\odot$ carbon white dwarf with a
$0.05~M_\odot$ He layer.  We seed the He layer with 1\% (by mass) of
\isot{N}{14}.  The white dwarf is isothermal and the temperature ramps
up at the base of the He layer and is then isentropic until the
surface.  Figure~\ref{fig:subch_initial_model} shows the structure of
the initial model.  We use 2d axisymmetric coordinates with a domain
size of $1.024\times 10^9~\mathrm{cm} \times 2.048\times
10^9~\mathrm{cm}$, a base grid of $256 \times 512$ zones and two
levels of refinement (both jumping by $2\times$), giving a fine-grid
resolution of $10~\mathrm{km}$.  Gravity is modeled as a monopole.
Our choice of initial perturbation has the detonation propagate inward
into the carbon white dwarf.  We run the simulation for 0.1 s---this
is enough time for the carbon detonation to be well developed.

@@

An initial perturbation is placed on the symmetry axis of the form:
\begin{equation}
  T = T_0 \left [ 1 + X(\isotm{He}{4}) f (1 + \tanh(2 - r1)) \right ]
\end{equation}
where
\begin{equation}
  r_1 = \left [ x^2 + (y - r_0)^2 \right ]^{1/2} / \lambda
\end{equation}
and
\begin{equation}
  r_0 = r_\mathrm{pert} + r_\mathrm{base}
\end{equation}
Here, $r_\mathrm{base}$ is the radius at which the helium layer begins
and $r_\mathrm{pert}$ is the distance above the base to put the perturbation.  We
choose $r_\mathrm{pert} = 100~\mathrm{km}$.  The temperature is pertubed above
the initial model value, denoted as $T_0$ here.  The amplitude of the perturbation
is $f = 12$ and the scale of the perturbation is $\lambda = 200~\mathrm{km}$.

We do a suite of runs with both Strang and the simplified-SDC integration, using
CFL numbers of 0.1, 0.2, and 0.4 and reaction network tolerances of $\atolspec =
\rtolspec = 10^{-5}$ and $\atolener = \rtolener = 10^{-65$.  We also run the CFL
= 0.2 case with a tighter tolerance, using  $\atolspec = 10^{-8}$.  For the step
rejection part of VODE, we use $f = 4$ and $\eta = 4$.

Figure \ref{fig:subch_strang_cfl08_I} shows the temperature, mean
molecular weight, and nuclear energy generation rate for the strang
CFL = 0.8 simulation ({\tt strang\_subch2\_cfl0.8}).  We see that the
mean molecular weight appears mottled behind the inward propagating
detonation in the white dwarf.  Comparing to the energy generation
rate, we see that region is characterized by large energy releases of
alternating signs checkerboarding thoughout the region.  This is
characteristic of the nucleosynthesis struggling to come into
equilibrium at high temperatures when the reverse rates can be
important.  In Figure~\ref{fig:subch_strang_cfl08_II} we look at some
important mass fractions for the same simulation.  Again we see that
the abundances are not smooth behind the inward detonation.

\begin{figure}
\centering
\plotone{strang_subch2_cfl0_8_state}
\caption{\label{fig:subch_strang_cfl08_I}}
\end{figure}

\begin{figure}
\centering
\plotone{strang_subch2_cfl0_8_mass_fractions}
\caption{\label{fig:subch_strang_cfl08_II}}
\end{figure}

We next look at the simplified-SDC version of this same case, {\tt
  sdc\_subch2\_cfl0.8}.  Figure~\ref{fig:subch_sdc_cfl08_I} shows the
temperature, mean molecular weight, and energy generation rate.
Compared to Figure~\ref{fig:subch_strang_cfl08_I}, we see that the
composition and energy generation rates are smooth, without any of the
checkerboarding that plagued the Strang simulation.  The compositions
plot (figure~\ref{fig:subch_sdc_cfl08_II}) shows the same behavior.
This suggests that the simplified-SDC algorithm has an easier time
attaining the correct equilibrium in these conditions.

\begin{figure}
\centering
\plotone{sdc_subch2_cfl0_8_state}
\caption{\label{fig:subch_sdc_cfl08_I}}
\end{figure}

\begin{figure}
\centering
\plotone{sdc_subch2_cfl0_8_mass_fractions}
\caption{\label{fig:subch_sdc_cfl08_II}}
\end{figure}

@@

We next look at how the structure seen in
Figures~\ref{fig:subch_strang_cfl08_I} and \ref{fig:subch_sdc_cfl08_I}
vary with CFL number and network tolerance.  We focus just on the
region behind the inward propagating shock.
Figures~\ref{fig:strang_enuc_summary} and
\ref{fig:strang_abar_summary} show the nuclear energy generation rate
and mean molecular weight for all of the Strang runs.  We see that
changing the CFL number does little to affect the state behind the
shock---it continues to take on the mottled appearance characteristic
of not reaching a proper equilibrium.  For the {\tt mixedtol} case, we
see that the state is greatly improved.  We were unable to run the
{\tt tol1.e-6} case for Strang, as the code aborted with too many
subcycles.  While reducing the tolerances does help, it greatly
increases the cost of the simulation.

\begin{figure}
\plotone{strang_enuc_summary_plot.pdf}
\caption{\label{fig:strang_enuc_summary} Comparison of Strang runs showing
the nuclear energy generation rate behind the inward propagating
shock.}
\end{figure}

\begin{figure}
\plotone{strang_abar_summary_plot.pdf}
\caption{\label{fig:strang_abar_summary} Comparison of Strang runs showing
the mean molecular weight behind the inward propagating
shock.}
\end{figure}

@@

\begin{figure}
\plotone{sdc_enuc_summary_plot.pdf}
\caption{\label{fig:sdc_enuc_summary} Comparison of simplified-SDC runs showing
the nuclear energy generation rate behind the inward propagating
shock.}
\end{figure}

\begin{figure}
\plotone{sdc_abar_summary_plot.pdf}
\caption{\label{fig:sdc_abar_summary} Comparison of simplified-SDC runs showing
the mean molecular weight behind the inward propagating
shock.}
\end{figure}

@@

We presented a simplified spectral deferred corrections scheme for coupling
hydrodynamics and reactions.  We demonstrated that for a moderate-sized reaction
network, the simplified-SDC method

We have not focused on performance or load-balancing the reactions in
the tests that we ran, but these limited runs show that the
simplified-SDC method runs as fast (or faster) than the Strang method
for the same tolerances and CFL, despite having to do the advection
twice.  We would expect similar behavior for any simulations that are
dominated by the cost of reactions.  If the results above a robust across
different problems and networks, then this suggests that we can run 
the SDC simulation with larger CFL and/or less strict reaction tolerances,
which would translate into a large performance boost.

@@

\acknowledgements \castro\ is freely available at
\url{http://github.com/AMReX-Astro/Castro}.  All of the code and
problem setups used here are available in the git repo.  The work at
Stony Brook was supported by DOE/Office of Nuclear Physics grant
DE-FG02-87ER40317.  This material is based upon work supported by the
U.S. Department of Energy, Office of Science, Office of Advanced
Scientific Computing Research and Office of Nuclear Physics,
Scientific Discovery through Advanced Computing (SciDAC) program under
Award Number DE-SC0017955.  This research was supported by the
Exascale Computing Project (17-SC-20-SC), a collaborative effort of
the U.S. Department of Energy Office of Science and the National
Nuclear Security Administration.

@@

The final Jacobian is found by multiplying these two:
\begin{equation}
\renewcommand{\arraystretch}{1.5}
\frac{\partial \Rb}{\partial \Uc^\prime} = \left (
  \begin{array}{ccc}
    \frac{\partial \omegadot_\alpha}{\partial X_\alpha} - \frac{e_{X_\alpha}}{c_v} \frac{\partial \omegadot_\alpha}{\partial T} &
    \frac{\partial \omegadot_\alpha}{\partial X_\beta} - \frac{e_{X_\beta}}{c_v} \frac{\partial \omegadot_\alpha}{\partial T} &
    \frac{1}{c_v} \frac{\partial \omegadot_\alpha}{\partial T} \\
     %
    \frac{\partial \omegadot_\beta}{\partial X_\alpha} - \frac{e_{X_\alpha}}{c_v} \frac{\partial \omegadot_\beta}{\partial T} &
    \frac{\partial \omegadot_\beta}{\partial X_\beta} - \frac{e_{X_\beta}}{c_v} \frac{\partial \omegadot_\beta}{\partial T} &
    \frac{1}{c_v} \frac{\partial \omegadot_\beta}{\partial T} \\
     %
     \frac{\partial \Sdot}{\partial X_\alpha} -  \frac{e_{X_\alpha}}{c_v} \frac{\partial \Sdot}{\partial T} &
     \frac{\partial \Sdot}{\partial X_\beta} -  \frac{e_{X_\beta}}{c_v} \frac{\partial \Sdot}{\partial T} &
     \frac{1}{c_v} \frac{\partial \Sdot}{\partial T} \\
  \end{array}
  \right )
\renewcommand{\arraystretch}{1}
\end{equation}
We note that the form of these entries is the same as one would arrive at if you start with the rates expressed as
$\omega_k(\rho, T(\rho, X_j, e), X_j)$ and recognize that constant $e$ implies that
\begin{equation}
\left . \frac{\partial T}{\partial X_k} \right |_{\rho, e} = - \frac{e_{X_k}}{c_v}
\end{equation}

While VODE can compute the entire Jacobian, ${\bf J}$ numerically via
differencing, we found that does not give reliable results.  Instead,
we compute compute the derivatives with respect to $X_k$ and $T$
one-sided differencing, following the algorithm in \cite{lsode} to
minimize numerical noise.  We then use the equation of state to
compute $e_{X_k}$ and $c_v$ and construct the entries of the final
Jacobian.

>>>>

    \item construct the advective update terms, $\Advt{\Uc}^{n+1/2,(k)}_{i}$,
      by first ignoring the hydrodynamics sources,
      \begin{equation}
        \Advt{\Uc}^{n+1/2,(k)}_{i} =
          - \frac{\Fb^\xv(\Uc^{n+1/2,(k)}_{i+1/2}) - \Fb^\xv(\Uc^{n+1/2,(k)}_{i-1/2})}{\Delta x}
      \end{equation}
      for the general case, and
      \begin{align}
        \Advsst{\rho e}^{n+1/2,(k)}_{i} =
          &- \frac{F\left((\rho e)^{n+1/2,(k)}_{i+1/2}\right) - F\left((\rho e)^{n+1/2,(k)}_{i-1/2}\right)}{\Delta x} \nonumber \\
          &- \left (\frac{p_{i+1/2}^{n+1/2,(k)} + p_{i-1/2}^{n+1/2,(k)}}{2} \right )\frac{u_{i+1/2}^{n+1/2,(k)} - u_{i-1/2}^{n+1/2,(k)}}{\Delta x}
      \end{align}
      for $(\rho e)$.

@@

The initial model is generated following the procedure described in
\citep{subchandra}.  We use a $1.1~M_\odot$ carbon white dwarf with a
$0.05~M_\odot$ He layer.  We seed the He layer with 1\% (by mass) of
\isot{N}{14}.  The white dwarf is isothermal and the temperature ramps
up at the base of the He layer and is then isentropic until the
surface.  Figure~\ref{fig:subch_initial_model} shows the structure of
the initial model.  We use 2d axisymmetric coordinates with a domain
size of $1.024\times 10^9~\mathrm{cm} \times 2.048\times
10^9~\mathrm{cm}$, a base grid of $256 \times 512$ zones and a single
level of refinement (jumping by $2\times$), giving a fine-grid
resolution of $20~\mathrm{km}$.  Gravity is modeled as a monopole.
Our choice of initial perturbation has the detonation propagate inward
into the carbon white dwarf.  We run the simulation for 0.1 s---this
is enough time for the carbon detonation to be well developed.

@@

An initial perturbation is placed on the symmetry axis of the form:
\begin{equation}
  T = T_0 \left [ 1 + X(\isotm{He}{4}) f (1 + \tanh(2 - r_1)) \right ]
\end{equation}
where
\begin{equation}
  r_1 = \left [ x^2 + (y - r_0)^2 \right ]^{1/2} / \lambda
\end{equation}
and
\begin{equation}
  r_0 = r_\mathrm{pert} + r_\mathrm{base}
\end{equation}
Here, $r_\mathrm{base}$ is the radius at which the helium layer begins
and $r_\mathrm{pert}$ is the distance above the base to put the perturbation.  We
choose $r_\mathrm{pert} = 100~\mathrm{km}$.  The temperature is pertubed above
the initial model value, denoted as $T_0$ here.  The amplitude of the perturbation
is $f = 12$ and the scale of the perturbation is $\lambda = 200~\mathrm{km}$.

We do a suite of runs with both Strang and the simplified-SDC integration, using
CFL numbers of 0.1, 0.2, and 0.4 and reaction network tolerances of $\atolspec =
\rtolspec = 10^{-5}$ and $\atolener = \rtolener = 10^{-5}$.  We also run the CFL
= 0.2 case with a tighter tolerance, using  $\atolspec = 10^{-8}$.  For the step
rejection part of VODE, we use $f = 4$ and $\eta = 4$.

Figure \ref{fig:subch_strang_cfl02} shows the temperature, mean
molecular weight, and nuclear energy generation rate for the strang
CFL = 0.82 simulation ({\tt strang\_subch2\_cfl0.2}).  We see that the
mean molecular weight appears mottled behind the inward propagating
detonation in the white dwarf.  Comparing to the energy generation
rate, we see that region is characterized by large energy releases of
alternating signs checkerboarding thoughout the region.  This is
characteristic of the nucleosynthesis struggling to come into
equilibrium at high temperatures when the reverse rates can be
important.

\begin{figure}
\centering
\plotone{strang_subch2_cfl0.2_tols_slice}
\caption{\label{fig:subch_strang_cfl02}}
\end{figure}

We next look at the simplified-SDC version of this same case, {\tt
  sdc\_subch2\_cfl0.2}.  Figure~\ref{fig:subch_sdc_cfl02} shows the
temperature, mean molecular weight, and energy generation rate.
Compared to Figure~\ref{fig:subch_strang_cfl02}, we see that the
composition and energy generation rates are smooth, without any of the
checkerboarding that plagued the Strang simulation.
This suggests that the simplified-SDC algorithm has an easier time
attaining the correct equilibrium in these conditions.

\begin{figure}
\centering
\plotone{sdc_subch2_cfl0.2_tols_slice}
\caption{\label{fig:subch_sdc_cfl02}}
\end{figure}

@@

We next look at how the structure seen in
Figures~\ref{fig:subch_strang_cfl02} and \ref{fig:subch_sdc_cfl02}
vary with CFL number and network tolerance.  We focus just on the
region behind the inward propagating shock.
Figures~\ref{fig:strang_enuc_summary} and
\ref{fig:strang_abar_summary} show the nuclear energy generation rate
and mean molecular weight for all of the Strang runs.  We see that
changing the CFL number does little to affect the state behind the
shock---it continues to take on the mottled appearance characteristic
of not reaching a proper equilibrium.  For the case with tight tolerances, we
see that the state is greatly improved.

\begin{figure}
\plotone{strang_enuc_summary_plot}
\caption{\label{fig:strang_enuc_summary} Comparison of Strang runs showing
the nuclear energy generation rate behind the inward propagating
shock.}
\end{figure}

\begin{figure}
\plotone{strang_abar_summary_plot}
\caption{\label{fig:strang_abar_summary} Comparison of Strang runs showing
the mean molecular weight behind the inward propagating
shock.}
\end{figure}

@@

\begin{figure}
\plotone{sdc_enuc_summary_plot}
\caption{\label{fig:sdc_enuc_summary} Comparison of simplified-SDC runs showing
the nuclear energy generation rate behind the inward propagating
shock.}
\end{figure}

\begin{figure}
\plotone{sdc_abar_summary_plot}
\caption{\label{fig:sdc_abar_summary} Comparison of simplified-SDC runs showing
the mean molecular weight behind the inward propagating
shock.}
\end{figure}

@@

We presented a simplified spectral deferred corrections scheme for coupling
hydrodynamics and reactions.  We demonstrated that for a moderate-sized reaction
network, the simplified-SDC method compares well against Strang for the double
detonation problem---getting a better solution at looser reaction tolerances,
and consistent results across a range of CFL numbers.  The benefits of this new
integration scheme are likely problem and network dependent, and we will explore
it on other science problemsin future papers.

We have not focused on performance or load-balancing the reactions in
the tests that we ran, but for the simulations shown here,
the simplified-SDC method runs as fast (or faster) than the Strang method
for the same tolerances and CFL, despite having to do the advection and reactions
twice.  We would expect similar behavior for any simulations that are
dominated by the cost of reactions.  If the results above a robust across
different problems and networks, then this suggests that we can run
the SDC simulation with larger CFL and/or less strict reaction tolerances,
which would translate into a large performance boost.

@@

\acknowledgements \castro\ is freely available at
\url{http://github.com/AMReX-Astro/Castro}.  All of the code and problem setups
used here are available in the git repo.  The double detonation problem is in
{\tt Castro/Exec/science/subchandra}.  The work at Stony Brook was supported by
DOE/Office of Nuclear Physics grant DE-FG02-87ER40317.  This material is based
upon work supported by the U.S. Department of Energy, Office of Science, Office
of Advanced Scientific Computing Research and Office of Nuclear Physics,
Scientific Discovery through Advanced Computing (SciDAC) program under Award
Number DE-SC0017955.  This research was supported by the Exascale Computing
Project (17-SC-20-SC), a collaborative effort of the U.S. Department of Energy
Office of Science and the National Nuclear Security Administration.  This
research used resources of the National Energy Research Scientific Computing
Center (NERSC), a U.S. Department of Energy Office of Science User Facility
located at Lawrence Berkeley National Laboratory, operated under Contract No.
DE-AC02-05CH11231 using NERSC award NP-ERCAP0020354.

@@

The final Jacobian is found by multiplying these two:
\begin{equation}
\renewcommand{\arraystretch}{1.5}
\frac{\partial \Rb}{\partial \Uc^\prime} = \left (
  \begin{array}{ccc}
    \frac{\partial \omegadot_\alpha}{\partial X_\alpha} - \frac{e_{X_\alpha}}{c_v} \frac{\partial \omegadot_\alpha}{\partial T} &
    \frac{\partial \omegadot_\alpha}{\partial X_\beta} - \frac{e_{X_\beta}}{c_v} \frac{\partial \omegadot_\alpha}{\partial T} &
    \frac{1}{c_v} \frac{\partial \omegadot_\alpha}{\partial T} \\
     %
    \frac{\partial \omegadot_\beta}{\partial X_\alpha} - \frac{e_{X_\alpha}}{c_v} \frac{\partial \omegadot_\beta}{\partial T} &
    \frac{\partial \omegadot_\beta}{\partial X_\beta} - \frac{e_{X_\beta}}{c_v} \frac{\partial \omegadot_\beta}{\partial T} &
    \frac{1}{c_v} \frac{\partial \omegadot_\beta}{\partial T} \\
     %
     \frac{\partial \Sdot}{\partial X_\alpha} -  \frac{e_{X_\alpha}}{c_v} \frac{\partial \Sdot}{\partial T} &
     \frac{\partial \Sdot}{\partial X_\beta} -  \frac{e_{X_\beta}}{c_v} \frac{\partial \Sdot}{\partial T} &
     \frac{1}{c_v} \frac{\partial \Sdot}{\partial T} \\
  \end{array}
  \right )
\renewcommand{\arraystretch}{1}
\end{equation}
We note that the form of these entries is the same as one would arrive at if you start with the rates expressed as
$\omega_k(\rho, T(\rho, X_j, e), X_j)$ and recognize that constant $e$ implies that
\begin{equation}
\left . \frac{\partial T}{\partial X_k} \right |_{\rho, e,X_{j,j\ne k}} = - \frac{e_{X_k}}{c_v}
\end{equation}

Finally we note that one implication of this formulation is that the Jacobian is no longer sparse.  Future
work will explore alternate formulations and approximate Jacobians.

% While VODE can compute the entire Jacobian, ${\bf J}$ numerically via
% differencing, we found that does not give reliable results.  Instead,
% we compute compute the derivatives with respect to $X_k$ and $T$
% one-sided differencing, following the algorithm in \cite{lsode} to
% minimize numerical noise.  We then use the equation of state to
% compute $e_{X_k}$ and $c_v$ and construct the entries of the final
% Jacobian.
